---
title: "eDNA Bayesian joint species distribution model"
author: "Noémie Coulon"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

<!--
This script documents the workflow used to prepare eDNA data, fit a spatially
blocked joint species distribution model (JSDM), evaluate marginal predictions,
and generate posterior community realisations for best-of-N analyses.

The sample-level eDNA data and environmental covariates required to run this
workflow are not distributed with the repository. The associated 50-km
aggregated dataset is provided separately for data sharing and reuse, but does
not allow the analyses in this script to be rerun. This repository is therefore
intended as transparent analytical documentation rather than as an executable
reproducibility package.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Reproducible eDNA analysis

## Dependencies and configuration

Install packages once outside this script (ideally with `renv`). The code does
not install packages automatically, which makes executions reproducible.

```{r packages}
required_packages <- c("boral", "blockCV", "dplyr", "readr", "sf", "terra")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages)) {
  stop("Install before running: ", paste(missing_packages, collapse = ", "), call. = FALSE)
}

invisible(lapply(required_packages, library, character.only = TRUE))

data_dir <- "data/ecolet"
output_dir <- "output/ecolet/boral_cv"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

sequence_file <- file.path(data_dir, "seq.csv")
metadata_file <- file.path(data_dir, "MED_metadatas_2018-2025.csv")
bathymetry_file <- file.path(data_dir, "gebco_2026_n44.5_s38.5_w0.0_e12.5.nc")
required_files <- c(sequence_file, metadata_file, bathymetry_file)

if (!all(file.exists(required_files))) {
  stop("Missing input file(s): ", paste(required_files[!file.exists(required_files)], collapse = ", "), call. = FALSE)
}

folds <- seq_len(5)
n_posterior_draws <- 50L
presence_threshold <- 0.5
continuous_variables <- c("depth", "volume_sampled", "temperature")
set.seed(123)
```

## Helper functions

```{r functions}
extract_or_nearest <- function(raster, longitude, latitude) {
  
  point <- terra::vect(matrix(c(longitude, latitude), ncol = 2), crs = terra::crs(raster))
  
  value <- terra::extract(raster, point, cells = TRUE)[[2]]
  
  if (!is.na(value)) return(list(value = value, lon = longitude, lat = latitude, replaced = FALSE))

  valid_cells <- which(!is.na(terra::values(raster)))
  
  if (!length(valid_cells)) return(list(value = NA_real_, lon = NA_real_, lat = NA_real_, replaced = FALSE))
  
  valid_coordinates <- terra::xyFromCell(raster, valid_cells)
  
  nearest_cell <- valid_cells[which.min(terra::distance(terra::vect(valid_coordinates, crs = terra::crs(raster)), point))]
  
  nearest_coordinates <- terra::xyFromCell(raster, nearest_cell)
  
  list(value = terra::values(raster)[nearest_cell], lon = nearest_coordinates[1], lat = nearest_coordinates[2], replaced = TRUE)
  
}

micro_f1 <- function(observed, predicted) {
  true_positive <- sum(observed == 1 & predicted == 1)
  false_positive <- sum(observed == 0 & predicted == 1)
  false_negative <- sum(observed == 1 & predicted == 0)
  denominator <- 2 * true_positive + false_positive + false_negative
  if (denominator == 0) NA_real_ else 2 * true_positive / denominator
}

community_f1 <- function(observed, predicted) {
  true_positive <- rowSums(observed == 1 & predicted == 1)
  false_positive <- rowSums(observed == 0 & predicted == 1)
  false_negative <- rowSums(observed == 1 & predicted == 0)
  denominator <- 2 * true_positive + false_positive + false_negative
  ifelse(denominator == 0, NA_real_, 2 * true_positive / denominator)
}

simulate_posterior_probability <- function(fit, mcmc_samples, posterior_row, X_new, ranef_new) {
  n_samples <- nrow(X_new)
  n_species <- fit$p
  n_latent_variables <- fit$num.lv
  parameter_names <- colnames(mcmc_samples)
  posterior_draw <- mcmc_samples[posterior_row, , drop = TRUE]

  latent_index <- grep("^lv\\.coefs\\[", parameter_names)
  if (length(latent_index) != n_species * (n_latent_variables + 1)) stop("Unexpected latent-variable coefficients.")
  latent_coefficients <- matrix(as.numeric(posterior_draw[latent_index]), nrow = n_species, ncol = n_latent_variables + 1)

  fixed_index <- grep("^X\\.coefs\\[", parameter_names)
  if (length(fixed_index) != n_species * fit$num.X) stop("Unexpected fixed-effect coefficients.")
  beta <- matrix(as.numeric(posterior_draw[fixed_index]), nrow = n_species, ncol = fit$num.X)

  linear_predictor <- matrix(latent_coefficients[, 1], nrow = n_samples, ncol = n_species, byrow = TRUE) + X_new %*% t(beta)
  latent_scores <- matrix(rnorm(n_samples * n_latent_variables), nrow = n_samples)
  linear_predictor <- linear_predictor + latent_scores %*% t(latent_coefficients[, -1, drop = FALSE])

  for (random_effect in seq_len(ncol(ranef_new))) {
    sd_index <- grep(paste0("^ranef\\.sigma\\.ID", random_effect, "\\["), parameter_names)
    if (length(sd_index) != n_species) stop("Unexpected random-effect standard deviations.")
    group_id <- as.integer(ranef_new[, random_effect])
    if (!identical(sort(unique(group_id)), seq_len(max(group_id)))) stop("Random-effect IDs must be consecutive.")
    random_coefficients <- matrix(
      rnorm(max(group_id) * n_species, sd = rep(as.numeric(posterior_draw[sd_index]), each = max(group_id))),
      nrow = max(group_id), ncol = n_species)
    linear_predictor <- linear_predictor + random_coefficients[group_id, , drop = FALSE]
  }
  probability <- pnorm(linear_predictor)
  colnames(probability) <- colnames(fit$y)
  probability
}
```

# Data preparation

## Read and harmonise eDNA data

```{r read-data}
sequence_data <- readr::read_csv(sequence_file, show_col_types = FALSE)
metadata <- readr::read_delim(metadata_file, delim = ";", trim_ws = TRUE, show_col_types = FALSE)

sequence_data <- sequence_data |>
  dplyr::select(spygen_code, dplyr::where(~ is.numeric(.x) && any(.x != 0, na.rm = TRUE))) |>
  dplyr::filter(spygen_code %in% metadata$spygen_code)

sampling_data <- metadata |>
  dplyr::filter(spygen_code %in% sequence_data$spygen_code) |>
  dplyr::transmute(
    spygen_code,
    date = as.Date(date),
    lat = (latitude_start_DD + latitude_end_DD) / 2,
    lon = (longitude_start_DD + longitude_end_DD) / 2,
    mpa_fully = dplyr::recode(as.character(Protection), "1" = "0", "2" = "1"),
    volume_sampled = dplyr::coalesce(flow_sensor_volume, estimated_volume),
    depth = NA_real_)
```

## Extract environmental covariates

For coastal points where a raster value is unavailable, the closest valid cell
is used and reported in the console.

```{r extract-environment}
bathymetry <- terra::rast(bathymetry_file)
for (i in seq_len(nrow(sampling_data))) {
  
  extraction <- extract_or_nearest(bathymetry, sampling_data$lon[i], sampling_data$lat[i])
  sampling_data$depth[i] <- abs(extraction$value)
  
  if (extraction$replaced) message("Nearest bathymetry cell used for ", sampling_data$spygen_code[i])
  
}

sst_files <- list.files(data_dir, pattern = "SST", full.names = TRUE)
if (!length(sst_files)) stop("No SST raster files were found in ", data_dir, call. = FALSE)
sst <- do.call(c, lapply(sst_files, terra::rast))
sst_dates <- as.Date(terra::time(sst))

sampling_data$temperature <- NA_real_
sampling_data$temperature_lon <- NA_real_
sampling_data$temperature_lat <- NA_real_

for (i in seq_len(nrow(sampling_data))) {
  date_index <- which(sst_dates == sampling_data$date[i])
  
  if (!length(date_index)) {
    warning("No SST layer for ", sampling_data$spygen_code[i], " on ", sampling_data$date[i])
    next
  }
  
  extraction <- extract_or_nearest(sst[[date_index[1]]], sampling_data$lon[i], sampling_data$lat[i])
  sampling_data$temperature[i] <- extraction$value - 273.15 # Kelvin to Celsius
  sampling_data$temperature_lon[i] <- extraction$lon
  sampling_data$temperature_lat[i] <- extraction$lat
  if (extraction$replaced) message("Nearest SST cell used for ", sampling_data$spygen_code[i])
  
}

sampling_data <- sampling_data |>
  dplyr::filter(dplyr::between(lat, 39, 44), depth > 0, temperature > 0)
readr::write_csv(sampling_data, file.path(data_dir, "environmental_data.csv"))
```

## Build model matrices

```{r model-matrices}
presence_absence <- sequence_data |>
  dplyr::filter(spygen_code %in% sampling_data$spygen_code) |>
  dplyr::mutate(dplyr::across(-spygen_code, ~ as.integer(.x > 0)))

metadata_model <- sampling_data |>
  dplyr::filter(spygen_code %in% presence_absence$spygen_code, !is.na(volume_sampled)) |>
  dplyr::arrange(match(spygen_code, presence_absence$spygen_code)) |>
  dplyr::mutate(
    date_numeric = as.numeric(date - min(date, na.rm = TRUE)),
    mpa_fully = factor(mpa_fully),
    log_volume = log1p(volume_sampled))

presence_absence_model <- presence_absence |>
  dplyr::filter(spygen_code %in% metadata_model$spygen_code) |>
  dplyr::arrange(match(spygen_code, metadata_model$spygen_code))
stopifnot(identical(metadata_model$spygen_code, presence_absence_model$spygen_code))

Y <- as.matrix(dplyr::select(presence_absence_model, -spygen_code))
storage.mode(Y) <- "numeric"
stopifnot(all(Y %in% c(0, 1)))

readr::write_csv(metadata_model, file.path(data_dir, "metadata_model.csv"))
readr::write_csv(presence_absence_model, file.path(data_dir, "eDNA_presence_absence.csv"))
saveRDS(Y, file.path(data_dir, "Y.rds"))
```

# Spatial cross-validation

```{r spatial-folds}
metadata_model <- metadata_model |>
  dplyr::mutate(
    station = interaction(round(lon, 6), round(lat, 6), drop = TRUE),
    date_group = as.character(date))
metadata_sf <- sf::st_as_sf(metadata_model, 
                            coords = c("lon", "lat"), crs = 4326, remove = FALSE)

set.seed(123)
spatial_cv <- blockCV::cv_spatial(
  x = metadata_sf, column = NULL, k = length(folds), size = 50000,
  selection = "random", iteration = 100, biomod2 = FALSE, progress = TRUE)

metadata_model$spatial_fold <- spatial_cv$folds_ids
stopifnot(setequal(unique(metadata_model$spatial_fold), folds))
readr::write_csv(metadata_model, file.path(data_dir, "metadata_model.csv"))
```

## Fit and predict each fold

The fitting chunk is not evaluated when knitting because it is computationally
demanding. Run it interactively or set `eval = TRUE` on a configured machine.

```{r fit-folds, eval=FALSE}
species_keep_by_fold <- sapply(folds, function(fold) {
  training_rows <- metadata_model$spatial_fold != fold
  n_presence <- colSums(Y[training_rows, , drop = FALSE])
  n_presence >= 5 & n_presence <= sum(training_rows) - 5
})

Y_cv <- Y[, apply(species_keep_by_fold, 1, all), drop = FALSE]

for (fold in folds) {
  message("Fitting fold ", fold, " of ", length(folds))
  set.seed(123 + fold)
  test_id <- which(metadata_model$spatial_fold == fold)
  train_id <- which(metadata_model$spatial_fold != fold)
  metadata_train <- metadata_model[train_id, , drop = FALSE]
  metadata_test <- metadata_model[test_id, , drop = FALSE]
  Y_train <- Y_cv[train_id, , drop = FALSE]
  Y_test <- Y_cv[test_id, , drop = FALSE]

  x_center <- vapply(metadata_train[continuous_variables], mean, numeric(1), na.rm = TRUE)
  x_scale <- vapply(metadata_train[continuous_variables], sd, numeric(1), na.rm = TRUE)
  stopifnot(all(is.finite(x_scale)), all(x_scale > 0))
  
  make_X <- function(data) {
    X <- scale(as.matrix(data[continuous_variables]), 
               center = x_center, scale = x_scale)
    X <- cbind(X, 
               AMP = as.numeric(as.character(data$mpa_fully)))
    storage.mode(X) <- "numeric"; X
  }
  
  make_ranef <- function(data) {
    random_effects <- cbind(station = as.integer(factor(data$station)), 
                            date = as.integer(factor(data$date_group)))
    storage.mode(random_effects) <- "integer"; random_effects
  }
  
  X_train <- make_X(metadata_train); X_test <- make_X(metadata_test)
  ranef_train <- make_ranef(metadata_train); ranef_test <- make_ranef(metadata_test)
  stopifnot(!anyNA(X_train), !anyNA(X_test), !anyNA(ranef_train), !anyNA(ranef_test))

  fit <- boral::boral(
    y = Y_train, 
    X = X_train, 
    family = "binomial", 
    ranef.ids = ranef_train,
    lv.control = list(num.lv = 2, type = "independent"),
    save.model = TRUE,
    mcmc.control = list(n.burnin = 100,
                        n.iteration = 500,
                        n.thin = 5, 
                        seed = 123 + fold))
  
  prediction <- predict(fit, newX = X_test, newranef.ids = ranef_test,
    predict.type = "marginal", scale = "response", lv.mc = 100, return.alllinpred = FALSE)
  
  probability <- prediction$linpred
  stopifnot(identical(dim(probability), dim(Y_test)))

  saveRDS(fit, file.path(output_dir, paste0("fit_fold", fold, ".rds")))
  
  saveRDS(list(test_id = test_id, observed = Y_test, probability = probability,
    predicted_pa = as.integer(probability >= presence_threshold), x_center = x_center,
    x_scale = x_scale, ranef_test = ranef_test),
    file.path(output_dir, paste0("predictions_fold", fold, ".rds")))
}
```

# Out-of-fold evaluation

```{r evaluate}
prediction_files <- file.path(output_dir, paste0("predictions_fold", folds, ".rds"))

fit_files <- file.path(output_dir, paste0("fit_fold", folds, ".rds"))

if (!all(file.exists(c(prediction_files, fit_files)))) stop("Fit all folds before evaluation.", call. = FALSE)

predictions_by_fold <- lapply(prediction_files, readRDS)

all_test_ids <- unlist(lapply(predictions_by_fold, `[[`, "test_id"))

stopifnot(length(all_test_ids) == nrow(metadata_model), 
          anyDuplicated(all_test_ids) == 0)

species_names <- colnames(predictions_by_fold[[1]]$observed)

oof_probability <- matrix(NA_real_, 
                          nrow(metadata_model), 
                          length(species_names), 
                          dimnames = list(metadata_model$spygen_code, 
                                          species_names))

oof_observed <- matrix(NA_integer_, 
                       nrow(metadata_model), 
                       length(species_names), 
                       dimnames = dimnames(oof_probability))

fold_summary <- dplyr::bind_rows(lapply(seq_along(folds), function(i) {
  
  prediction <- predictions_by_fold[[i]]; test_id <- prediction$test_id
  
  stopifnot(identical(colnames(prediction$observed), species_names))
  
  oof_probability[test_id, ] <<- prediction$probability
  
  oof_observed[test_id, ] <<- prediction$observed
  
  data.frame(fold = folds[i], n_train = nrow(metadata_model) - length(test_id), n_test = length(test_id),
             
    F1_micro = micro_f1(prediction$observed, prediction$predicted_pa))
}))

stopifnot(!anyNA(oof_probability), !anyNA(oof_observed))

oof_predicted_pa <- as.integer(oof_probability >= presence_threshold)

marginal_oof_results <- list(
  probability = oof_probability, observed = oof_observed, predicted_pa = oof_predicted_pa,
  F1_by_sample = community_f1(oof_observed, oof_predicted_pa),
  F1_micro = micro_f1(oof_observed, oof_predicted_pa), fold_summary = fold_summary
)
saveRDS(marginal_oof_results, file.path(output_dir, "marginal_oof_results.rds"))
fold_summary
```

# Posterior community realisations

Each realisation samples one joint posterior draw, new latent scores, and new
random-effect coefficients. It is thus a plausible community configuration,
not a repeated thresholding of marginal probabilities.

```{r posterior-draws, eval=FALSE}
probability_tables_by_fold <- vector("list", length(folds))
names(probability_tables_by_fold) <- as.character(folds)

for (fold in folds) {
  
  message("Generating posterior draws for fold ", fold)
  
  fit <- readRDS(file.path(output_dir, paste0("fit_fold", fold, ".rds")))
  
  prediction_info <- readRDS(file.path(output_dir, paste0("predictions_fold", fold, ".rds")))
  
  test_id <- prediction_info$test_id
  metadata_test <- metadata_model[test_id, , drop = FALSE]
  Y_test <- prediction_info$observed
  X_test <- scale(as.matrix(metadata_test[continuous_variables]), center = prediction_info$x_center, scale = prediction_info$x_scale)
  X_test <- cbind(X_test, AMP = as.numeric(as.character(metadata_test$mpa_fully)))
  storage.mode(X_test) <- "numeric"
  ranef_test <- as.matrix(prediction_info$ranef_test); storage.mode(ranef_test) <- "integer"
  
  stopifnot(identical(colnames(Y_test), colnames(fit$y)), identical(colnames(X_test), colnames(fit$X)),
    identical(colnames(ranef_test), colnames(fit$ranef.ids)))

  mcmc_samples <- as.matrix(boral::get.mcmcsamples(fit))
  set.seed(1000 + fold)
  posterior_rows <- sample(seq_len(nrow(mcmc_samples)), 
                           n_posterior_draws, 
                           replace = nrow(mcmc_samples) < n_posterior_draws)
  
  probability_array <- simplify2array(lapply(posterior_rows, function(row) {
    
    simulate_posterior_probability(fit, mcmc_samples, row, X_test, ranef_test)
  }))
  
  dimnames(probability_array) <- list(spygen_code = metadata_test$spygen_code,
                                      species = colnames(Y_test), 
                                      draw = paste0("draw_", seq_len(n_posterior_draws)))

  probability_matrix <- matrix(aperm(probability_array, c(3, 1, 2)),
    nrow = nrow(Y_test) * n_posterior_draws,
    ncol = ncol(Y_test),
    dimnames = list(NULL, colnames(Y_test)))
  
  probability_table <- dplyr::bind_cols(
    data.frame(row_id = rep(test_id, 
                            each = n_posterior_draws),
               spygen_code = rep(metadata_test$spygen_code,
                                 each = n_posterior_draws),
               fold = fold,
               draw = rep(seq_len(n_posterior_draws),
                          times = nrow(Y_test))),
    as.data.frame(probability_matrix, check.names = FALSE))
  
  saveRDS(list(test_id = test_id,
               spygen_code = metadata_test$spygen_code,
               posterior_rows = posterior_rows,
               probability_array = probability_array),
          file.path(output_dir, 
                    paste0("probability_array_fold", fold, "_", n_posterior_draws, ".rds")))
  
  saveRDS(probability_table,
          file.path(output_dir, paste0("probability_table_fold", fold, "_", n_posterior_draws, ".rds")))
  
  probability_tables_by_fold[[as.character(fold)]] <- probability_table
}

probability_draws <- dplyr::bind_rows(probability_tables_by_fold) |>
  dplyr::arrange(row_id, draw)

stopifnot(nrow(probability_draws) == nrow(metadata_model) * n_posterior_draws)

readr::write_csv(probability_draws, file.path(output_dir, paste0("probability_draws_", n_posterior_draws, ".csv")))
saveRDS(probability_draws, file.path(output_dir, paste0("probability_draws_", n_posterior_draws, ".rds")))
```

# Export shareable data

The following export is a public release using 50-km grid-centroid
coordinates, but retains sampling dates and covariates.

```{r export-data}
metadata_50km <- metadata_model |>
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) |>
  sf::st_transform(3035)
grid_sf <- sf::st_sf(grid_id = seq_along(sf::st_make_grid(metadata_50km, 
                                                          cellsize = 50000, 
                                                          square = TRUE)),
  geometry = sf::st_make_grid(metadata_50km, 
                              cellsize = 50000, 
                              square = TRUE))

metadata_50km <- sf::st_join(metadata_50km, grid_sf)

centroids <- sf::st_transform(sf::st_centroid(grid_sf), 4326)
centroid_coordinates <- sf::st_coordinates(centroids)
grid_sf$lon_50km <- centroid_coordinates[, "X"]
grid_sf$lat_50km <- centroid_coordinates[, "Y"]

shared_data <- metadata_50km |>
  sf::st_drop_geometry() |>
  dplyr::left_join(sf::st_drop_geometry(grid_sf) |>
    dplyr::select(grid_id, lon_50km, lat_50km), 
    by = "grid_id") |>
  dplyr::select(spygen_code, 
                date, 
                depth, 
                mpa_fully, 
                volume_sampled, 
                temperature,
                spatial_fold,
                lon_50km,
                lat_50km) |>
  dplyr::left_join(presence_absence_model, by = "spygen_code")

readr::write_csv(shared_data, file.path(data_dir, "Coulon_et_al_2027_eDNA_data.csv"))
```
